\documentclass[aspectratio=169]{beamer}

\input{../preamble/beamer_styles.tex}

% -----------------------------
% Title info
% -----------------------------
\title[MMAE 450]{Nonlinear Systems in Two Dimensions}
\subtitle{Two-Term Linearization in \(\,x\) and \(\,y\)}
\author{MMAE 450}
\date{}

\begin{document}

% -------------------------------------------------
\begin{frame}
  \titlepage
\end{frame}

% -------------------------------------------------
\begin{frame}{Nonlinear Systems of Equations}
Many engineering problems lead to systems of nonlinear equations:
\[
\mathbf{F}(\mathbf{x}) = \mathbf{0},
\quad
\mathbf{x} =
\begin{bmatrix}
x \\ y
\end{bmatrix}.
\]

\vspace{0.75em}
Examples include:
\begin{itemize}
  \item nonlinear material behavior,
  \item geometric nonlinearity,
  \item nonlinear boundary conditions.
\end{itemize}

\vspace{0.75em}
To solve these systems, we rely on \emph{local linearization}.
\end{frame}

% -------------------------------------------------
\begin{frame}{A Two-Dimensional Nonlinear System}
Consider a system of two nonlinear equations:
\[
\begin{aligned}
F_1(x,y) &= 0, \\
F_2(x,y) &= 0.
\end{aligned}
\]

\vspace{0.75em}
We seek a solution \((x,y)\) that satisfies both equations simultaneously.

\vspace{0.75em}
In general, these equations cannot be solved analytically.
\end{frame}

% -------------------------------------------------
\begin{frame}{Local Linearization About a Point}
Let \((x_0,y_0)\) be a current approximation to the solution.

\vspace{0.75em}
We approximate the nonlinear system near this point using a first-order Taylor expansion.

\vspace{0.75em}
This is a direct generalization of the 1D Taylor expansion.
\end{frame}

% -------------------------------------------------
\begin{frame}{Two-Term Taylor Expansion (2D)}
Each component of \(\mathbf{F}\) is expanded about \((x_0,y_0)\):

\[
\begin{aligned}
F_1(x,y) \approx\;&
F_1(x_0,y_0)
+ \frac{\partial F_1}{\partial x}(x_0,y_0)(x-x_0)
+ \frac{\partial F_1}{\partial y}(x_0,y_0)(y-y_0), \\[0.5em]
F_2(x,y) \approx\;&
F_2(x_0,y_0)
+ \frac{\partial F_2}{\partial x}(x_0,y_0)(x-x_0)
+ \frac{\partial F_2}{\partial y}(x_0,y_0)(y-y_0).
\end{aligned}
\]

\vspace{0.5em}
Higher-order terms are neglected.
\end{frame}

% -------------------------------------------------
\begin{frame}{Linearized System in Matrix Form}
The linearized system can be written compactly as:
\[
\mathbf{F}(x,y)
\approx
\mathbf{F}(x_0,y_0)
+
\mathbf{J}(x_0,y_0)
\begin{bmatrix}
x - x_0 \\
y - y_0
\end{bmatrix},
\]

where \(\mathbf{J}\) is the Jacobian matrix:
\[
\mathbf{J} =
\begin{bmatrix}
\dfrac{\partial F_1}{\partial x} & \dfrac{\partial F_1}{\partial y} \\[0.8em]
\dfrac{\partial F_2}{\partial x} & \dfrac{\partial F_2}{\partial y}
\end{bmatrix}.
\]
\end{frame}

% -------------------------------------------------
\begin{frame}{What Does the Jacobian Represent?}
\begin{itemize}
  \item The Jacobian generalizes the concept of a derivative to multiple dimensions
  \item It describes how the system changes with respect to small changes in \((x,y)\)
  \item It provides the best linear approximation near the current point
\end{itemize}

\vspace{0.75em}
\begin{block}{Key Idea}
Nonlinear problems are solved by repeatedly solving linearized systems.
\end{block}
\end{frame}

% -------------------------------------------------
\begin{frame}{Why Linearization Matters}
\begin{itemize}
  \item Enables iterative solution methods (Newton-type methods)
  \item Converts nonlinear problems into a sequence of linear solves
  \item Forms the foundation of nonlinear finite element analysis
\end{itemize}

\vspace{0.75em}
\begin{block}{Looking Ahead}
Next, we will use this linearization to construct Newton's method
for solving nonlinear systems.
\end{block}
\end{frame}

\end{document}